C/O in Protoplanetary Disks: The Effect of Radial Drift and Viscous Accretion

### 2015-06-10 tex/draft/ms.tex
update

\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentclass[12pt, preprint]{aastex}
%\documentclass[apj]{emulateapj}

@@

\emgr{Background topics: importance of atmospheric chemistry in providing constraints on the formation of giant planets; C/O ratio as important signature of atmospheric chemistry; C/O ratios observationally determined are different from interstellar --- one explanation is the different abundance in gas and dust form of the main C and O carriers, H$_2$O, CO$_2$ and CO, between their respective snowlines (cite \"Oberg et al. 2011).}

@@

We describe our analytic model for the radial drift of solids in section \ref{sec:drift}. In section \ref{sec:disk}, we present our protoplanetary disk model, as well as our prescription for the gas and dust surface density evolution in an active disk. Finally, we summarize our ice desorption model in section \ref{sec:desorption}.

@@

\subsection{Evolutionary Disk Model}
\label{sec:disk}

@@

\emgr{Describe disk model --- both passive MMSN and active self-similar solution with $\alpha$ prescription for viscosity. Describe the model for evolving the surface density profile of planetesimals following Birnstiel et al. (2012). Describe temperature profile --- currently power-law, but may change. Describe desorption model and parameters following Hollenbach et al. (2009). Mention that the solids are perfect spheres composed of a single volatile. Perhaps this subsection needs to be split into subsubsections.}

\subsection{Relevant timescales}

@@

\emgr{Calculate timescale for radial drift following Chiang \& Youdin (2010). Calculate desorption timescale following Hollenbach et al. (2009). Estimate gas accretion timescale for a given $\alpha$. Important: mention that, for simplicity and illustrative purposes, these calculations are performed for a passive disk. Show plot with the timescales as a function of particle size at different snowlines to show the regime in which drift matters}.

\section{Snowline Locations}

@@

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/desorption_distance_vs_s.pdf}
%\vspace{-0.5in}
\caption{Desorption distance as a function of initial particle size, for the range of particles that desorb at a fixed distance regardless of initial location (see Figure \ref{fig:snowlines}). \textbf{Figure belongs to section 3.} \emgr{This is a figure that I'm not sure we should include, since it doesn't really provide that much useful inside. If we did include it though, there would be two panels both for the passive and the active disk. The caption is not complete since the active disk panel is not yet there.}}  %  (See text for a description of evolution to yet higher masses.) 
\label{fig:an_vs_actual}
\end{figure}

\emgr{Present the equation set that you are solving, $dr/dt=\dot{r}$, $ds/dt=...$. I don't think it is necessary to describe in detail the numerical method of solving the equations since it's pretty straightforward. Present the pretty rainbow plots side by side for passive and active disks, and highlight the differences. Insert the plots that shows that for the intermediate size particles, the desorption distance can be estimated analytically with good accuracy. Maybe also include the plot showing the desorption distance as a function of particle size, both for passive and active disk. Split into subsections?}

\section{Results for the C/O Ratio}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/C_O_ratio_active_disk_many.pdf}
%\vspace{-0.5in}
\caption{C/O ratio in gas (solid lines) and in dust (dashed lines) for a passive disk (left panel \emgr{currently non-existent}) and for an active disk (right panel \emgr{currently just a single plot}), and for the range of particle sizes that desorb at a fixed distance regardless of their initial location in the disk. The particle size increases from 0.001 cm to 700 cm in a rainbow color scheme. The horizontal dotted line represents the stellar value of 0.54. The black lines (\emgr{currently non-existent}) represent the C/O ratio in gas (solid black line) and dust (dashed black line) for a static disk. The snowline location moves inward as the particle size increases. \textbf{Figure belongs to section 4.}}
\label{fig:CO_ratio}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{lala}
%\vspace{-0.5in}
\caption{C/O ratios for different particle size distributions... \textbf{Figure belongs to section 4.} (\emgr{It will probably be a multi-panel plot, for different particle size distributions, for both passive and active disk, and perhaps at different times for the active disk. Not yet unsure how that figure will look like since I don't have the results yet, so I'll leave the figure caption open for now.})}
\label{fig:...}
\end{figure}

@@

\section{Discussion and Model Limitations}

>>>>

%\documentclass[12pt, preprint,numberedappendix]{emulateapj}
%\documentclass[12pt, preprint]{aastex}
\documentclass[apj]{emulateapj}

@@

%\emgr{Background topics: importance of atmospheric chemistry in providing constraints on the formation of giant planets; C/O ratio as important signature of atmospheric chemistry; C/O ratios observationally determined are different from interstellar --- one explanation is the different abundance in gas and dust form of the main C and O carriers, H$_2$O, CO$_2$ and CO, between their respective snowlines (cite \"Oberg et al. 2011).}

@@

We present our protoplanetary disk model for both a passive and an active disk in section \ref{sec:disk}. In section \ref{sec:drift}, we describe our analytic model for the radial drift of solids. We summarize our ice desorption model in section \ref{sec:desorption}. Finally, we discuss the relevant timescales for dynamical effects in the desorption process in section \ref{sec:timescales}.

\subsection{Disk Model}
\label{sec:disk}

We adopt a minimum mass solar nebula (MMSN) disk model for a passive disk similar to the prescription of \citet{chiang10}. The gas surface density and midplane temperature are
\begin{subeqnarray}
\label{eq:disk}
\Sigma_{\rm d}&=&2200\, (r/\text{AU})^{-1}\,\, \text{g cm}^{-2} \slabel{eq:disksigma}\\
T_{\rm d} &=& 120\, (r/\text{AU})^{-3/7} \,\,\text{K}, \slabel{eq:diskT}
\end{subeqnarray}
where $r$ is the semimajor axis. Based on some observations of protoplanetary disks \citep{andrews10}, our surface density profile, $\Sigma_{\rm d} \propto r^{-1}$, is flatter than that of \citet{chiang10}, i.e. $\Sigma_{\rm d} \propto r^{-3/2}$.

We model the active disk as a thin disk with an $\alpha$-viscosity prescription \citep{shakura73}:
\begin{equation}
\label{eq:nu}
\nu=\alpha c_{\rm d} H_{\rm d}.
\end{equation}
Here $\nu$ is the kinematic viscosity, $\alpha < 1$ is a dimensionless coefficient, and $c_{\rm d}$, $H_{\rm d}$ are the isothermal sound speed and disk scale height, respectively:
\begin{subeqnarray}
\label{eq:cdHd}
c_{\rm d} &=& \sqrt{\frac{k_{\rm B} T_{\rm d}}{\mu m_{\rm p}}} \slabel{eq:cd} \\
H_{\rm d}&=& \frac{c_{\rm d}}{\Omega_{\rm k}} \slabel{eq:Hd},
\end{subeqnarray}
where $k_{\rm B}$ is the Boltzmann constant, $\mu$ is the mean molecular weight of the gas, $m_{\rm p}$ is the proton mass, and $\Omega_{\rm k}$ is the Keplerian angular velocity, $\Omega_{\rm k} \equiv \sqrt{G M_*/r^3}$, with $G$ the gravitational constant and $M_*$ the stellar mass. The temperature profile for an active disk is assumed to be the same as for the passive disk and given by Equation (\ref{eq:diskT}). From Equations (\ref{eq:nu}) and (\ref{eq:cdHd}), the viscosity can thus be expressed as a power-law in radius, $\nu \propto r^{\gamma}$, with $\gamma=15/14 \approx 1$ for our choice of parameters. Following \citet{armitage10}, we define $R \equiv r/r_{\rm c}$ and $\nu_{\rm c} \equiv \nu(r_{\rm c})$, where $r_{\rm c}$ is a characteristic disk radius. The gas surface density is then given by the self-similar solution
\begin{equation}
\label{eq:Sigmaact}
\Sigma_{\rm d}(R, T) = \frac{M_{\rm d}}{2 \pi r_{\rm c}^2 R^{\gamma}} T^{(-5/2-\gamma)/(2-\gamma)} \exp{\Big[-\frac{R^{-(2-\gamma)}}{T}}\Big],
\end{equation}
where $M_{\rm d}$ is the total disk mass and
\begin{subeqnarray}
\label{eq:T}
T & \equiv & \frac{t}{t_{\rm c}} + 1 \\
t_{\rm c} & \equiv & \frac{1}{3(2-\gamma)} \frac{r_{\rm c}^2}{\nu_{\rm c}},
\end{subeqnarray}
where $t$ is time.

@@

Solid particles in a disk orbit their host star at the Keplerian velocity $v_{\rm k} \equiv \Omega_{\rm k} r$. The gas, however, experiences an additional pressure gradient, which causes it to rotate at sub-Keplerian velocity \citep{weidenschilling77}. Dust grains thus experience a headwind, which removes angular momentum, causing the solids to spiral inwards and fall onto the host star. Small particles are well-coupled to the gas, while large planetesimals are decoupled from the gas. From \citet{chiang10}, the extent of coupling is quantified by the dimensionless stopping time, $\tau_{\rm s} \equiv \Omega_{\rm k} t_{\rm s}$, where $t_{\rm s}$ is
\begin{equation}
\label{eq:ts}
t_{\rm s}= \left\{
\begin{array}{l l}
\rho_{\rm s} s / (\rho_{\rm d} c_{\rm d}), & \quad s < 9 \lambda/4 \,\,\,\ \text{Epstein drag} \\
4 \rho_{\rm s} s^2 / (9 \rho_{\rm d} c_{\rm d} \lambda), & \quad s < 9 \lambda/4, \,\text{Re} \lesssim 1 \,\,\,\ \text{Stokes drag.}
\end{array} 
\right.
\end{equation}
Here $\rho_{\rm d}$ is the gas midplane density, $\rho_{\rm s}$ is the density of a solid particle, $s$ is the particle size, $\lambda$ is the mean free path, and Re is the Reynolds number.

For a passive disk, the radial drift velocity can be approximated analytically as
\begin{equation}
\label{eq:rdotpas}
\dot{r} \approx -2 \eta \Omega_{\rm k} r \Big(\frac{\tau_{\rm s}}{1+\tau_{\rm s}^2}\Big),
\end{equation}
where
\begin{equation}
\label{eq:eta}
\eta \equiv - \frac{\partial P_{\rm d}/\partial \ln r}{2 \rho_{\rm d} v_{\rm k}^2} \approx \frac{c_{\rm d}^2}{2 v_k^2}
\end{equation}
and $P_{\rm d} = \rho_{\rm d} c_{\rm d}^2$ is the disk midplane pressure.

For an active disk, the radial drift velocity has an additional term due to the radial movement of the gas, i.e.
\begin{equation}
\label{eq:rdotact}
\dot{r} \approx -2 \eta \Omega_{\rm k} r \Big(\frac{\tau_{\rm s}}{1+\tau_{\rm s}^2}\Big) + \frac{v_{\rm gas}}{1+\tau_{\rm s}^2},
\end{equation}
where $v_{\rm gas}$ is the radial gas accretion velocity and can be expressed as (e.g., \citealt{fkr02})
\begin{equation}
\label{eq:vgas}
v_{\rm gas}= - \frac{3}{\Sigma_{\rm d} \sqrt{r}} \frac{\partial}{\partial r}(\nu \Sigma_{\rm d} \sqrt{r}) 
\end{equation}
with $\Sigma_{\rm d}$ from Equation (\ref{eq:Sigmaact}).

@@

In order for a volatile species to thermally desorb, it has to overcome the binding energy that keeps it on the grain surface. Following \citet{hollenbach09}, the desorption rate per molecule for a species $x$ can be expressed as
\begin{equation}
\label{eq:Rdes}
R_{\rm{des}, x} = \nu_x \exp{(-E_x/T_{\rm grain})},
\end{equation}
where $E_x$ is the adsorption binding energy, $T_{\rm grain}$ is the grain temperature, and $\nu_x=1.6 \times 10^{11} \sqrt{(E_x/\mu_x)}$ is the molecule's vibrational frequency, with $\mu_x$ the mean molecular weight. We assume that the dust and gas have the same temperature in the disk midplane, hence $T_{\rm grain}=T_{\rm d}$. For H$_2$O, CO$_2$ and CO, the binding energies $E_x$ are assumed to be 5800 K, 2000 K and 850 K, respectively \citep{oberg11}. We use the desorption rate, $R_{\rm des}$, to estimate the desorption timescale for particles of different sizes as described in section \ref{sec:timescales}.

%\emgr{Describe disk model --- both passive MMSN and active self-similar solution with $\alpha$ prescription for viscosity. Describe the model for evolving the surface density profile of planetesimals following Birnstiel et al. (2012). Describe temperature profile --- currently power-law, but may change. Describe desorption model and parameters following Hollenbach et al. (2009). Mention that the solids are perfect spheres composed of a single volatile. Perhaps this subsection needs to be split into subsubsections.}

\subsection{Relevant timescales}
\label{sec:timescales}

We can estimate the extent to which radial drift and gas accretion affect desorption by comparing the timescales for desorption, drift and accretion, for solids of different sizes and compositions.

\textit{Desorption timescale.} We assume that the solid bodies are perfect spheres and are entirely composed of only one volatile species, i.e. either H$_2$O, CO$_2$ or CO \footnote{We discuss the validity of these simplifications in section \ref{sec:discussion}.}. The desorption timescale can then be estimated as
\begin{equation}
\label{eq:tdes}
t_{\rm des}=\frac{\rho_{\rm s}}{3 \mu_x m_{\rm p}} \frac{s}{N_x R_{\rm des, x}},
\end{equation}
where $N_x \approx 10^{15}$ sites cm$^{-2}$ is the number of adsorption sites of volatile $x$ per cm$^2$ \citep{hollenbach09}.

\textit{Radial drift timescale.} To order of magnitude, the radial drift timescale can be estimated as 
\begin{equation}
\label{eq:tdrift}
t_{\rm r, drift} \sim \frac{r}{\dot{r}},
\end{equation}
where $\dot{r}$ is given by Equation (\ref{eq:rdotpas}) for a passive disk and by Equation (\ref{eq:rdotact}) for an active disk.

\textit{Gas accretion timescale.} The timescale for gas accretion onto the central star for an active disk is (e.g., \citealt{armitage10})
\begin{equation}
\label{eq:tgas}
t_{\rm gas, acc} \sim \frac{r^2}{\nu} \sim \frac{1}{2 \alpha \eta \Omega_{\rm k}},
\end{equation}
with the latter expression derived from Equations (\ref{eq:nu}) and (\ref{eq:eta}).

For simplicity purposes, we calculate the radial drift timescale, $t_{\rm r, drift}$, for a passive disk in this section, but most of our conclusions hold true for an active disk as well. Figure \ref{fig:timescales} shows $t_{\rm des}$, $t_{\rm r, drift}$ and $t_{\rm gas, acc}$ as a function of particle size at three different locations in the disk, for $\rho_{\rm s}=2$ g cm$^{-3}$ and $\alpha=0.01$. As expected, micron-sized particles desorb on very short timescales of $\sim 1-1000$ years and their radial drift timescale exceeds the typical disk lifetime of a few Myr by several orders of magnitude. Thus for small particles in a passive disk, the snowline locations and the C/O ratio are the same as for a static disk (see Figure 1 from \citealt{oberg11}) \footnote{This is not true for an active disk, however, where gas accretion causes even micron-sized particles to drift significantly before desorbing, as we show in section \ref{sec:snowlines}.}. At the other extreme, kilometer-sized particles are unaffected by gas drag and have long desorption timescales ($\gg$1 Myr ), and the snowline locations and C/O ratio remain unchanged in this case as well. This is true for both passive and active disks, since large planetesimals are decoupled from the gas and hence unaffected by gas accretion onto the host star.

Of particular interest for our purposes is the particle size regime for which $t_{\rm r, drift} \lesssim t_{\rm des} \lesssim t_{\rm d}$ or $t_{\rm gas, acc} \lesssim t_{\rm des} \lesssim t_{\rm d}$, where $t_{\rm d}=3$ Myr is the disk lifetime. In these cases, radial drift or gas accretion (or both) are faster than thermal desorption. Particles of sizes that satisfy these requirements will drift significantly due to radial drift or gas accretion before desorbing, thus moving the H$_2$O, CO$_2$ and CO snowlines closer towards the central star and changing the C/O ratio throughout the disk. We quantify these effects in sections \ref{sec:snowlines} and \ref{sec:COratio}.

@@

%\emgr{Calculate timescale for radial drift following Chiang \& Youdin (2010). Calculate desorption timescale following Hollenbach et al. (2009). Estimate gas accretion timescale for a given $\alpha$. Important: mention that, for simplicity and illustrative purposes, these calculations are performed for a passive disk. Show plot with the timescales as a function of particle size at different snowlines to show the regime in which drift matters}.

\section{Snowline Locations}
\label{sec:snowlines}

In this section we use the model described in section \ref{sec:model} to quantify the effects of radial drift (passive disk) or radial drift and gas accretion (active disk) on the snowline location, for dust particles of different sizes composed of either H$_2$O, CO$_2$ or CO. Specifically, we determine a particle's final location (i.e., where the particle either fully desorbs or remains at its initial size due to a long desorption timescale) as a function of its initial position in the disk, after the gas disk has dissipated. The disk lifetime, $t_{\rm d}$, is particularly relevant since this is the timescale on which giant planets form. The snowline locations at $t=t_{\rm d}$ throughout the protoplanetary disk determine the disk C/O ratio in gas at this time, and thus the C/O ratio in giant planet atmospheres that have formed \textit{in situ}.

As a fiducial model, we choose $M_*=M_{\odot}$, $M_{\rm d}=0.1 M_{\odot}$ and $r_{\rm c}=100$ AU. For each species $x$, we determine the final location in the disk of a particle of initial size $s_0$ by solving the following system of coupled differential equations:

\begin{subeqnarray}
\label{eq:ddt}
\frac{ds}{dt} &= & - \frac{3 \mu_x m_{\rm p}}{\rho_{\rm s}} N_x R_{\rm des, x}  \slabel{eq:dsdt} \\
\frac{dr}{dt} &=& \dot{r},
\end{subeqnarray}
where $R_{\rm des, x}$ is evaluated at $T=T_{\rm d}(r)$, and $\dot{r}$ is given by Equation (\ref{eq:rdotpas}) for a passive disk and Equation (\ref{eq:rdotact}) for an active disk. Equation (\ref{eq:dsdt}) can be derived straightforwardly from Equation (\ref{eq:tdes}). Our boundary conditions are $s(t_0)=s_0$, $r(t_0)=r_0$, and $s(t_{\rm d})=0$, where $t_0$ is the initial time at which we start the integration and $r_0$ is the initial location of the particle. We choose $t_0=1$ year, but our result is independent on the initial integration time as long as $t_0 \ll t_{\rm d}$.

Figure \ref{fig:snowlines} shows our results for H$_2$O, CO$_2$ and CO particles, and for both a passive and an active disk. The final distance of the dust particles is consistent with Figure \ref{fig:timescales}. Kilometer-sized bodies do not drift or desorb during the disk lifetime both for a passive and active disk. Micron- to mm-sized particles do not drift before desorbing in a passive disk, but they drift significantly in an active disk since they move at the same velocity as the accreting gas. For $0.5$ cm $\lesssim s_0 \lesssim$ 700 cm in a passive disk and $0.001$ cm $\lesssim s_0 \lesssim$ 700 cm in an active disk, we notice that particles of initial size $s_0$ desorb at a fixed distance $r_{\rm des}$ regardless of their original location in the disk. We show in section \ref{sec:COratio} that this result is essential in determining the C/O ratio throughout the disk for different particle sizes.

@@

Intuitively, this fixed $r_{\rm des}$ should be the location in the disk for which $t_{\rm r, drift} \sim t_{\rm des}$, given an initial particle size. Figure \ref{fig:an_vs_actual} shows the validity of this approximation for the range of particle sizes that desorb at a fixed distance in a passive and an active disk. We notice that the analytic approximation accurately reproduces the numerical result for most cases of interest.

@@

%\emgr{I didn't write the text for Figure 4 since we haven't yet decided if in the end we will include it in some form. We can }

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/desorption_distance_vs_s.pdf}
%\vspace{-0.5in}
\caption{Desorption distance as a function of initial particle size, for the range of particles that desorb at a fixed distance regardless of initial location (see Figure \ref{fig:snowlines}). \textbf{Figure belongs to section 3.} \emgr{This is a figure that I'm not sure we should include, since it doesn't really provide that much useful inside. If we did include it though, there would be two panels both for the passive and the active disk. The caption is not complete since the active disk panel is not yet there.}}  %  (See text for a description of evolution to yet higher masses.) 
\label{fig:r_vs_s}
\end{figure}

%\emgr{Present the equation set that you are solving, $dr/dt=\dot{r}$, $ds/dt=...$. I don't think it is necessary to describe in detail the numerical method of solving the equations since it's pretty straightforward. Present the pretty rainbow plots side by side for passive and active disks, and highlight the differences. Insert the plots that shows that for the intermediate size particles, the desorption distance can be estimated analytically with good accuracy. Maybe also include the plot showing the desorption distance as a function of particle size, both for passive and active disk. Split into subsections?}

\section{Results for the C/O Ratio}
\label{sec:COratio}

@@

\begin{figure}[h!]
\centering
\includegraphics[width=0.5\textwidth]{../../figs/C_O_ratio_active_disk_many.pdf}
%\vspace{-0.5in}
\caption{C/O ratio in gas (solid lines) and in dust (dashed lines) for a passive disk (left panel \emgr{currently non-existent}) and for an active disk (right panel \emgr{currently just a single plot}), and for the range of particle sizes that desorb at a fixed distance regardless of their initial location in the disk. The particle size increases from 0.001 cm to $\sim$700 cm in a rainbow color scheme. The horizontal dotted line represents the stellar value of 0.54. The black lines (\emgr{currently non-existent}) represent the C/O ratio in gas (solid black line) and dust (dashed black line) for a static disk. The snowline location moves inward as the particle size increases. \textbf{Figure belongs to section 4.}}
\label{fig:CO_ratio}
\end{figure}

%\begin{figure}[h!]
%\centering
%\includegraphics[width=0.5\textwidth]{lala}
%%\vspace{-0.5in}
%\caption{C/O ratios for different particle size distributions... \textbf{Figure belongs to section 4.} (\emgr{It will probably be a multi-panel plot, for different particle size distributions, for both passive and active disk, and perhaps at different times for the active disk. Not yet unsure how that figure will look like since I don't have the results yet, so I'll leave the figure caption open for now.})}
%\label{fig:...}
%\end{figure}

@@

\section{Discussion and Model Limitations}
\label{sec:discussion}
